---
title: "Expression analysis using the Human Protein Atlas"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
---

```{r, message=FALSE, warning=FALSE, include=FALSE}
library(knitr)
library(ggplot2)
library(tidyverse)
```

This notebook uses data available from the [Human Protein Atlas](https://www.proteinatlas.org/about) (HPA) to answer questions about the expression specificity of genes and proteins in tissues and cells.

Data from the HPA can be downloaded from the "[Downloadable data](https://www.proteinatlas.org/about/download)" page. For proteomics data download `normal_tissue.tsv.zip` and for subcellular locationalisation data download `subcellular_location.tsv.zip`.

```{bash eval=FALSE}
# download, unzip, and remove the zip tissue data
wget https://www.proteinatlas.org/download/normal_tissue.tsv.zip -O ../raw/normal_tissue.tsv.zip
unzip ../raw/normal_tissue.tsv.zip -d ../raw
rm ../raw/normal_tissue.tsv.zip

# download, unzip, and remove the subcellular location data
wget https://www.proteinatlas.org/download/subcellular_location.tsv.zip -O ../raw/subcellular_location.tsv.zip
unzip ../raw/subcellular_location.tsv.zip -d ../raw
rm ../raw/subcellular_location.tsv.zip
```

# Subcellular localisation

The subcellular localisation of proteins is based on immunofluorescently stained cells. The tab-delimited file contains 12,073 proteins with information on the gene and the localisation.

```{r read_subcellular}
subcellular <- read_tsv('../raw/subcellular_location.tsv')
dim(subcellular)
head(subcellular)
```

There are four levels of reliability: "Approved", "Enhanced", "Supported", and "Uncertain".

```{r table_reliability}
table(subcellular$Reliability)
```

Subcellular compartment information is stored in four columns, with the column name reflecting the level of reliability: enhanced locations (`Enhanced`), supported locations (`Supported`), Approved locations (`Approved`), and uncertain locations (`Uncertain`).

A protein can have more than one approved location.

```{r location_info}
subcellular %>%
  filter(Reliability == "Approved") %>%
  filter(grepl('Cytosol', Approved)) %>%
  select(Gene, Approved) %>%
  head()
```

# Tissue specificity

HPA also provides cell and tissue expression profiles for proteins based on immunohistochemisty using tissue microarrays.

```{r load_tissue_spec}
tissue <- read_tsv('../raw/normal_tissue.tsv')
dim(tissue)
head(tissue)
```

The data is qualitative and the expression level is stored in the `Level` column. There are four expression levels: "High", "Medium", "Low", and "Not detected".

```{r tissue_level}
table(tissue$Level)
```

The bar plot tallies the different levels of expression and colouring each observation by their reliability (the same four levels as the subcellular localisation data).
  
```{r barplot_level_reliability}
tissue %>%
  mutate(Level = factor(Level, levels=c('Not detected', 'Low', 'Medium', 'High'))) %>%
  mutate(Reliability = factor(Reliability, levels=c('Uncertain', 'Supported', 'Enhanced', 'Approved'))) %>%
  ggplot(aes(x=Level, fill=Reliability)) +
  geom_bar() +
  labs(title='Protein observations, by level and reliability', y='Num observations')
```

The `Not detected` level has the highest number of observations and most have an `Approved` or `Enhanced` reliability. The following definitions for the Reliability scores were copied from the [HPA documentation](https://www.proteinatlas.org/about/assays+annotation#ifr).

* `Enhanced` - One or more antibodies are enhanced validated and there is no contradicting data, for example literature describes experimental evidence for a different location.
* `Supported` - There is no enhanced validation of the used antibody, but the annotated localization is reported in literature.
* `Approved` - If the localization of the protein has not been previously described and was detected by only one antibody without additional antibody validation.
* `Uncertain` - If the antibody-staining pattern contradicts experimental data or no expression is detected on the RNA level.

